---
title: "RProjectpt2"
output: html_notebook
---

# import data
```{r}
comm <- read.csv("grassland_data/grassland.community.csv", header = TRUE, row.names = 1) #getting biodiversity data into R
#setwd("~/Desktop")
```

# Packages

```{r}
library(picante)
```

# Community data

```{r}
class(comm)
dim(comm) #gives us rows x columns of dataset
rownames(comm) #gives row names
head(colnames(comm)) #gives us column names
comm[1:5, 1:5] #gives us first 5 rows and columns
apply(comm, 1, sum) #total abundance in each sample
comm <- decostand(comm, method = "total") #get the relative abundance
apply(comm, 1, sum)
comm[1:5, 1:5] #new tranformed data
```

# Trait data

```{r}
traits <- read.csv("grassland_data/species.traits.csv", header = TRUE, row.names = 1) #new dataset with traits
head(traits)
pairs(traits) #plots correlation between each trait
traits <- log10(traits) #because the variables looked skewed, log transform the data
pairs(traits) #log transformed correlation
```

# Metadata

```{r}
metadata <- read.csv("grassland_data/plot.metadata.csv", header = TRUE, row.names = 1) #new dataset
head(metadata)
```

# Phylogeny

```{r}
phy <- read.tree("grassland_data/grassland.phylogeny.newick") #new dataset that is in Newwick format for phylogeny
class(phy) #this file contain phylo format
names(phy) #elements in phy

phy$tip.label[1:5] #list few species tip labels
Ntip(phy) #number of tips
plot(phy, cex = 0.5) #gives us the phylogeny tree
metadata$rel.moisture<-as.factor(metadata$rel.moisture) #change moisture level to factor
```

# Cleaning and matching data sets
```{r}
ls() #shows us all the data files we inputted
combined <- match.phylo.comm(phy, comm) #takes community object and phly and reports any species that are not present
phy <- combined$phy
comm <- combined$comm

combined <- match.phylo.data(phy, traits) #now do the same but with traits
phy <- combined$phy
traits <- combined$data
all.equal(rownames(comm), rownames(metadata)) #checks to see if it works
```

# Visualizing and summarizing biodiversity data

## Community richness and diversity

```{r}
boxplot(specnumber(comm) ~ metadata$rel.moisture, ylab = "# of species") #boxplot comparing number of species in different moisture

model<-lm(specnumber(comm) ~ metadata$rel.moisture)
anova(model) #do anova instead of t-test because there are more than two samples

plot(specaccum(comm), xlab = "# of samples", ylab = "# of species") #accumulion curve of all samples: accumulion curve: graph that shows how many new species are found as more sampling are done
```

# Multivariate community analysis

## Hierarchical clustering

```{r}
# calculate Bray-Curtis distance among samples
comm.bc.dist <- vegdist(comm, method = "bray") #abundance weight measure of how similar each sample are to each other
# cluster communities using average-linkage algorithm
comm.bc.clust <- hclust(comm.bc.dist, method = "average") #hcust is used to create hierarchical clusters 
# plot cluster diagram
plot(comm.bc.clust, ylab = "Bray-Curtis dissimilarity") #way to plot the difference of each sample and create a phylogenic tree
```

## Ordination

```{r}
comm.bc.mds <- metaMDS(comm, dist = "bray") #nonmetric multidemensional scaling and uses stable solution scaling
stressplot(comm.bc.mds) #creates a stress plot used to show stress distribution and analyze behavior of a component

ordiplot(comm.bc.mds, display = "sites", type = "text") #plot distance on a x and y plot: ordiplot plots are low dimentsional ordination spaces for community data

ordipointlabel(comm.bc.mds) #plots ordination plots with points and text label to the points

#ordination plots that compares distance of the e types of moisture levels spatially
mds.fig <- ordiplot(comm.bc.mds, type = "none")
# plot just the samples, colour by habitat, pch=19 means plot a circle
points(mds.fig, "sites", pch = 19, col = "green", select = metadata$rel.moisture == 
    "1")
points(mds.fig, "sites", pch = 19, col = "blue", select = metadata$rel.moisture == 
    "2")
points(mds.fig, "sites", pch = 19, col = "blue", select = metadata$rel.moisture == 
    "3")
ordiellipse(comm.bc.mds, metadata$rel.moisture, conf = 0.95, label = TRUE) # add confidence ellipses around habitat types
ordicluster(comm.bc.mds, comm.bc.clust, col = "gray") # overlay the cluster results we calculated earlier
ordisurf(comm.bc.mds, comm[, "Sphaeralcea_coccinea"], bubble = TRUE, main = "Sphaeralcea coccinea abundance", 
    cex = 3) #plot Sphaeralcea abundance on a ordination diagram by fitting smooth surfaces for continuous variables 
```

## Adding environmental and trait data to ordinations

```{r}
ordiplot(comm.bc.mds)
plot(envfit(comm.bc.mds, metadata[, 3:6])) #fir com.bc.mds poitns with tbe the metadata and plotting metadta points on it
```

# Trait evolution

## Phylogenetic signal

```{r}
apply(traits, 2, Kcalc, phy) #applying Kcalc function to each column: Kcal=K statistic of plylogenetic signal: K>1 means strong phylogenetic signal and conservation traits
multiPhylosignal(traits, multi2di(phy)) #this produces K values ot p values
```

## Visualizing trait evolution

```{r}
#plots plylogenetic tree of including the leaf area data
plot(phy, direction = "up", show.tip.label = FALSE, show.node.label = TRUE, 
    cex = 0.7)
# Plot leaf area on the phylogeny. cex argument scales symbol size by trait
tiplabels(pch = 19, col = "black", cex = 3 * (traits[, "LeafArea"]/max(traits[, 
    "LeafArea"])))
```

## Phylogenetic analysis of trait relationships

```{r}
#for non-independent due to phylogenetic signal using generalized least squares which works like ANOVA or linear model
root.gls <- gls(RootTissueDens ~ SRL, data = traits) #compare root tissue and root length
anova(root.gls)

#include effects of phylogeny to model
root.pgls <- gls(RootTissueDens ~ SRL, correlation = corBrownian(value = 1, 
    phy), data = traits)
anova(root.pgls)

#plots root tissue density vs root length with GLS line of best fit and phylogenetic GLS fit
plot(RootTissueDens ~ SRL, data = traits, xlab = "SRL (specific root length)", 
    ylab = "Root tissue density")
# add model fit lines - coef is the model fit coefficients, lwd increases
# line width
abline(coef(root.gls), lwd = 2, col = "black")
abline(coef(root.pgls), lwd = 2, col = "red")
legend("bottomleft", legend = c("GLS fit", "Phylogenetic GLS fit"), lwd = 2, 
    col = c("black", "red"))
```

# Phylogenetic and trait diversity

## Phylogenetic diversity

```{r}
#using Faith PD (phylogenetic diversity) to calculate total branch length including all speces. Outputs PD values and species richness (SR)
comm.pd <- pd(comm, phy)
head(comm.pd)

boxplot(comm.pd$PD ~ metadata$rel.moisture, xlab = "Mositure", ylab = "Faith's PD") #plot FD with types of moisture

model2<-lm(comm.pd$PD ~ metadata$rel.moisture) #test using ANOVA on PD among moisture levels 
anova(model2)

plot(comm.pd$PD ~ comm.pd$SR, xlab = "Species richness", ylab = "Faith's PD") #plot faith's PD with species richness, there is a high correlation
```
## MPD, MNTD, SESMPD and SESMNTD

```{r}
#make a distance matrix using cophenetic which is used to find a correlation that measures how well a dendrogram represents the data
phy.dist <- cophenetic(phy)

comm.sesmpd <- ses.mpd(comm, phy.dist, null.model = "richness", abundance.weighted = FALSE, 
    runs = 999)
head(comm.sesmpd)# calculate ses.mpd: standardized effect size of mean pairwise distance

comm.sesmntd <- ses.mntd(comm, phy.dist, null.model = "richness", abundance.weighted = FALSE, 
    runs = 999)
head(comm.sesmntd) #calculate ses.mntd which is standardized effect size of mean nearest taxon distance to communities

plot(comm.sesmpd$mpd.obs.z ~ metadata$rel.moisture, xlab = "Moisture", ylab = "SES(MPD)")
abline(h = 0, col = "gray") #comparing the ses.mpd across all moisture types

model3<- lm(comm.sesmpd$mpd.obs.z ~ metadata$rel.moisture) #comparing ses across moisture types using ANOVA
anova(model3)

plot(comm.sesmntd$mntd.obs.z ~ metadata$rel.moisture, xlab = "Moisture", ylab = "SES(MNTD)")
abline(h = 0, col = "gray") #comparing the ses.mntd across all moisture types

model4<-lm(comm.sesmntd$mntd.obs.z ~ metadata$rel.moisture) #comparing sesmntd with mositure types
anova(model4)

plot(phy, show.tip.label = FALSE, main = "Fescue community fes-K-11")
tiplabels(tip = which(phy$tip.label %in% colnames(comm)[comm["fes-K-11", ] > 
    0]), pch = 19) #plot species present in fescue community

plot(phy, show.tip.label = FALSE, main = "Fescue community mix-H-23")
tiplabels(tip = which(phy$tip.label %in% colnames(comm)[comm["mix-H-23", ] > 
    0]), pch = 19) #plot species present in mixedgrass community
```

# Trait diversity

```{r}
trait.dist <- as.matrix(dist(scale(traits), method = "euclidean"))  #calcualting trait distance using Eulidean and produce a full distance matrix

comm.sesmpd.traits <- ses.mpd(comm, trait.dist, null.model = "richness", abundance.weighted = FALSE, 
    runs = 999) #using euclidean distance matrix to calculate trait ses.mpd

metadata$rel.moisture<-as.factor(metadata$rel.moisture)
plot(comm.sesmpd.traits$mpd.obs.z ~ metadata$rel.moisture, xlab = "moisture", ylab = "Trait SES(MPD)")
abline(h = 0, col = "gray") #plotting each
```

# Plylogenetic beta-dversity

```{r}
comm.mntd.dist <- comdistnt(comm, phy.dist, abundance.weighted = TRUE) #calculate  MNTD of phylogentic using comdistnt function: MNTD: mean nearest taxon distance

comm.mntd.traits.dist <- comdistnt(comm, trait.dist, abundance.weighted = TRUE) #calculate functional trait suing MNTD

mantel(comm.bc.dist, comm.mntd.dist) #calculate mantel: mental statistic is a correlation between 2 matrices (phylogenetic MNTD and functional trait MNTD)

mantel(comm.bc.dist, comm.mntd.traits.dist) #mantel correlation toxonomic bray-curtis and trait MNTD
```

# Phylogeny/trait-based ordinations

```{r}
#creates ordination of samples based on the MNTD matrices and phylogenetic distances
comm.mntd.mds <- monoMDS(comm.mntd.dist) #MNTD

mds.fig <- ordiplot(comm.mntd.mds, type = "none")
points(mds.fig, "sites", pch = 19, col = "green", select = metadata$rel.moisture == 
    "1")
points(mds.fig, "sites", pch = 19, col = "blue", select = metadata$rel.moisture == 
    "2")
points(mds.fig, "sites", pch = 19, col = "blue", select = metadata$rel.moisture == 
    "3")
# add confidence ellipses around habitat types
ordiellipse(comm.mntd.mds, metadata$rel.moisture, conf = 0.95, label = TRUE)

#using trait distance
comm.mntd.traits.mds <- monoMDS(comm.mntd.traits.dist)
# set up the plotting area but don't plot anything yet
mds.fig <- ordiplot(comm.mntd.traits.mds, type = "none")
# plot just the samples, colour by habitat, pch=19 means plot a circle
points(mds.fig, "sites", pch = 19, col = "green", select = metadata$rel.moisture == 
    "1")
points(mds.fig, "sites", pch = 19, col = "blue", select = metadata$rel.moisture == 
    "2")
points(mds.fig, "sites", pch = 19, col = "blue", select = metadata$rel.moisture == 
    "3")
# add confidence ellipses around habitat types
ordiellipse(comm.mntd.traits.mds, metadata$rel.moisture, conf = 0.95, label = TRUE)
```

# Testing for multivariate differences among groups

```{r}
#measure relationship of dissimilarity measures using MANOVA using adonis function
adonis(comm.bc.dist ~ rel.moisture, data = metadata) #comparing how moisture can explain taxonomic 

adonis(comm.mntd.dist ~ rel.moisture, data = metadata) #comparing how moisture can explain phylogenetic 

adonis(comm.mntd.traits.dist ~ rel.moisture, data = metadata) #comparing how moisture can explain  dissimilarity
```
